Notes on observational processes in epidemic models

### 2019-07-21 ms/ms.tex
stuff

Mechanistic analyses of an epidemic time series allow us to infer the underlying 
transmission mechanism, estimate biologically relevant parameters, and 
predict the course of an epidemic. In order to make a precise and accurate inference, 
disease modelers often focus on developing more realistic process models. For example,
England and Wales measles time series from the prevaccination era have been analyzed
using many different models -- these model account for time-varying transmission
rates \citep{fine1982measles}, realistic age structure \citep{schenzle1984age},
metapopulation structure \citep{xia2004measles}, continuous-time infection process 
\citep{cauchemez2008likelihood}, and extra-demographic variability \citep{he2009plug}.

Despite the amount of effort put into developing better process models, disease 
modelers often neglect details of the observation process associated with disease
case reports. Most disease models assume that new cases are reported
instantaneously when an individual is infected or shows symptom; one exception is
the model by \citep{he2009plug}, which assumes that infections are counted upon
recovery (because diagnosed cases are put to bed rest and are effectively 
no longer infectious).

@@

The Susceptible-Infected-Removed (SIR) model describes how disease spreads in a
homogeneous population:
\begin{equation}
\begin{aligned}
\frac{dS}{dt} &= - \beta S \frac{I}{N}\\
\frac{dI}{dt} &= \beta S \frac{I}{N} - \gamma I\\
\frac{dR}{dt} &= \gamma I,
\end{aligned}
\end{equation}
where $\beta$ is the contact rate, $\gamma$ is the recovery rate, and $N = S + I + R$ is
the total population size. Incidence measured upon infection ($i_1(t)$) can be
defined by the integral:
\begin{equation}
i_1(t) = \int_{t}^{t + \Delta t} \beta S \frac{I}{N} dt,
\end{equation}
where $\Delta t$ is the reporting time step.
Alternatively, we can keep track of cumulative incidence, $C$, by adding a 
new state variable $dC/dt = \beta S I/N$ and take the difference between 
two consecutive reporting periods: $i_1(t) = C(t+\delta t) - C(t)$. Likewise, 
incidence measured upon recovery ($i_2(t)$) can be defined by the integral:
\begin{equation}
i_2(t) = \int_{t}^{t + \Delta t} \gamma I dt,
\end{equation}
or by the difference in cumulative number of recovered cases: $i_2(t) = R(t + \Delta t) - R(t)$.
Finally, we assume that the reported incidence follows a negative binomial 
distribution with mean $\rho i_1(t)$ or $\rho i_2(t)$, where $\rho$ is the 
reporting rate, and a dispersion parameter $\theta$.

@@

Although incidence measured at two different periods are qualitatively very similar,

>>>>

Mechanistic analyses of epidemic time series allow us to infer the underlying 
transmission mechanism, estimate biologically relevant parameters, and 
predict the course of an epidemic \citep{breto2009time}. In order to make a 
precise and accurate inference, disease modelers often focus on developing
more realistic process models. For example, the same London measles time 
series from the prevaccination era has been analyzed using multiple models -- 
these models account for time-varying transmission rates \citep{fine1982measles}, 
realistic age structure \citep{schenzle1984age},
metapopulation structure \citep{xia2004measles}, continuous-time infection process 
\citep{cauchemez2008likelihood}, and extra-demographic variability \citep{he2009plug}.

Despite the amount of effort put into developing \emph{better} process models, disease 
modelers often neglect details of the observation process associated with disease
case reports. Most disease models assume that new cases are reported
instantaneously when an individual is infected (e.g., \cite{martinez2016differential, 
kennedy2018modeling, pons2018serotype}) or develops symptom; one exception is
the model by \citep{he2009plug}, which assumes that infections are counted upon
recovery (because diagnosed cases are put to bed rest and are effectively 
no longer infectious).

@@

The Susceptible-Infected-Removed (SIR) model describes how disease spreads in a
homogeneous population:
\begin{equation}
\begin{aligned}
\frac{dS}{dt} &= - \beta S \frac{I}{N}\\
\frac{dI}{dt} &= \beta S \frac{I}{N} - \gamma I\\
\frac{dR}{dt} &= \gamma I,
\end{aligned}
\end{equation}
where $\beta$ is the contact rate, $\gamma$ is the recovery rate, and $N = S + I + R$ is
the total population size. Incidence measured upon infection ($i_1(t)$) can be
defined by the integral:
\begin{equation}
i_1(t) = \int_{t}^{t + \Delta t} \beta S \frac{I}{N} dt,
\end{equation}
where $\Delta t$ is the reporting time step.
Alternatively, we can keep track of cumulative incidence, $C$, by adding a 
new state variable $dC/dt = \beta S I/N$ and take the difference between 
two consecutive reporting periods: $i_1(t) = C(t+\delta t) - C(t)$. Likewise, 
incidence measured upon recovery ($i_2(t)$) can be defined by the integral:
\begin{equation}
i_2(t) = \int_{t}^{t + \Delta t} \gamma I dt,
\end{equation}
or by the difference in cumulative number of recovered cases: $i_2(t) = R(t + \Delta t) - R(t)$.
Finally, we assume that the reported incidence follows a negative binomial 
distribution with mean $\rho i_1(t)$ or $\rho i_2(t)$, where $\rho$ is the 
reporting rate, and an over dispersion parameter $\theta$.

In this study, we focus on estimating 5 parameters: the basic reproductive
number $\mathcal R_0 = \beta/\gamma$, mean infectious period $1/\gamma$, 
reporting rate $\rho$, an over dispersion parameter $\theta$ and the initial
proportion of the infected individuals $i_0$. Initial proportion of
susceptible individuals is assumed to be $1 - i_0$. The total population
size $N$ is assumed to be known.

@@

Although the observed epidemic curves look very similar regardless of when incidence
is reported, fitting $i_1(t)$ and $i_2(t)$ to the same time series yields different
estimates (Figure ?).

To better understand how assumptions about timing of case reporting affect parameter
estimates of the SIR model, we simulate epidemic curves, measured upon both infection and
recovery, 100 times and fit both models to each time series (Figure ?).
